
---
title: "MPNST DMR report"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---


```{r include = FALSE}

# to knit document:
#library(rmarkdown)
#rmarkdown::render(
#  "/share/ScratchGeneral/jamtor/projects/MPNST_ctMet/scripts/2.filter_DMR.Rmd", 
#  output_file="/share/ScratchGeneral/jamtor/projects/MPNST_ctMet/results2/MPNST_DMR_report.html"
#)

subset_data <- TRUE
min_sample_per_probe <- 5
beta_cutoff <- 0.35
pval_cutoff <- as.character("1e-5")
background_cutoff <- 0.2
blood_cutoffs <- list(
  MPNST_hypermethylated = 0.2,
  MPNST_hypomethylated = 0.8
)
NF_cutoffs <- list(
  MPNST_hypermethylated = 0.2,
  MPNST_hypomethylated = 0.8
)
check_indiv_tissue <- as.logical("FALSE")
plot_min_sample_no <- 10

# define tissues:
tissue_key <- data.frame(
  code = c(
    "BLCA_non_malig", "blood", "BRCA_non_malig", 
    "CESC_non_malig", "CHOL_non_malig", "COAD_non_malig", 
    "ESCA_non_malig", "HNSC_non_malig", "KIRP_non_malig", 
    "LIHC_non_malig", "LUAD_non_malig", "LUSC_non_malig", 
    "mixed_non_malig", "MPNST", "NF", 
    "PAAD_non_malig", "PRAD_non_malig", "READ_non_malig", 
    "SARC_non_malig", "STAD_non_malig", "THCA_non_malig", 
    "THYM_non_malig"
  ),
  name = c(
    "bladder", "blood", "breast",
    "cervical", "bile duct", "colon",
    "esophagus", "head/neck", "kidney",
    "liver", "lung", "lung",
    "mixed", "MPNST", "NF", 
    "pancreas", "prostate", "rectum", 
    "sarcoma", "stomach", "thyroid", 
    "thymus"
  )
)

# define and create directories:
home_dir <- "/share/ScratchGeneral/jamtor/"
project_dir <- paste0(home_dir, "projects/MPNST_ctMet/")
script_dir <- paste0(project_dir, "scripts/")
func_dir <- paste0(script_dir, "functions/")
genome_dir <- paste0(project_dir, "genome/")
ref_dir <- paste0(project_dir, "refs/")

result_dir <- paste0(project_dir, "results2/")
Robject_in <- paste0(result_dir, "DMR/Rdata/")
out_path <- paste0(
  result_dir,
  "DMR/",
  "beta_", beta_cutoff, "_",
  "pval_", pval_cutoff, "/",
  "background_", background_cutoff, "/",
  "blood_hypo_", blood_cutoffs$MPNST_hypermethylated, "_",
  "blood_hyper_", blood_cutoffs$MPNST_hypomethylated, "/"
)

if (!is.na(NF_cutoffs[1])) {
  out_path <- paste0(
  	out_path, "NF_hypo_", NF_cutoffs$MPNST_hypermethylated,
  	"_NF_hyper_", NF_cutoffs$MPNST_hypomethylated, "/"
  )
}

if (subset_data) {
  result_dir <- paste0(result_dir, "subset/")
  out_path <- paste0(out_path, "subset/")
}

Robject_dir <- paste0(result_dir, "Rdata/")
plot_dir <- paste0(out_path, "plots/")
table_dir <- paste0(out_path, "tables/")

system(paste0("mkdir -p ", Robject_dir))
system(paste0("mkdir -p ", plot_dir))
system(paste0("mkdir -p ", table_dir))

print(
  paste0(
    "Output candidate plots are in ", 
    plot_dir
  )
)


####################################################################################
### 0. Load functions ###
####################################################################################

firstup <- function(x) {
  substr(x, 1, 1) <- toupper(substr(x, 1, 1))
  x
}



```

<br />
The following **450k methylation array** data was downloaded:


```{r echo = FALSE}

####################################################################################
### 1. Load and print sample and DMR info ###
####################################################################################

DMR_list <- readRDS(paste0(Robject_in, "MPNST_vs_normal_groups.Rdata"))
colnames(DMR_list$sample_summary) <- gsub(
  "_", " ", colnames(DMR_list$sample_summary)
)

library(knitr)

kable(DMR_list$sample_summary)

library(stringr)
merged_tissue_text <- firstup(paste(DMR_list$merged_tissues, collapse = ", "))


```

<br />
Original no. probes in non-malignant data = **`r DMR_list$non_malig_probes`**  \

Original no. probes in MPNST data = **`r DMR_list$MPNST_probes`**  \

**`r DMR_list$common_probes`** probes were present in both MPNST and non-malignant 
data and kept for analysis.  \ 

```{r eval = FALSE}

#<br />
#**`r merged_tissue_text`** data were merged as there were **< 5** samples per tissue, 
#creating a 'mixed tissue' group with `r DMR_list$mixed_no` samples.  \ 

```

```{r echo = FALSE}

# record DMR values:
DMR_record <- as.data.frame(
  rbind(
    sapply(DMR_list$MPNST_vs_normal, function(x) {
      nrow(
        x[x$status == "Hypermethylated in MPNST",]
      )
    }), 
    sapply(DMR_list$MPNST_vs_normal, function(x) {
      nrow(
        x[x$status == "Hypomethylated in MPNST",]
      )
    })
  )
)
rownames(DMR_record) <- c("unfiltered_hyper", "unfiltered_hypo")

```

#<br />
**`r print(DMR_record["unfiltered_hyper",]$all_non_malig)`** hypermethylated and
**`r print(DMR_record["unfiltered_hypo",]$all_non_malig)`** hypomethylated unfiltered 
candidates initially identified between the MPNST and non-malignant array data.


```{r, echo = FALSE}

####################################################################################
### 2. Filter by pval and beta difference ###
####################################################################################

print(
  paste0(
    "Filtering for canditates with at least ", 
    beta_cutoff, 
    " beta value difference and < ", pval_cutoff, "p-value..."
  )
)

# load function:
dget(paste0(func_dir, "DMR_filter.R"))
dget(paste0(func_dir, "record_vals.R"))

# filter:
DMR_initial <- lapply(
  MPNST_vs_normal,
  DMR_filter,
  beta_cutoff,
  as.numeric(pval_cutoff)
)

# record values:
DMR_record <- rbind(
  DMR_record,
  record_vals(DMR_initial, "beta_filt")
)

```

#<br />
Differentially methylated candidates with at least **`r print(beta_cutoff)`** difference 
in mean beta value and p-values < **`r print(pval_cutoff)`** were kept.
This left **`r print(DMR_record["",]$all_non_malig)`** hypermethylated and
**`r print(DMR_record["",]$all_non_malig)`** hypomethylated







